---
title: "`tidyDisasters` R Package: A harmonized disaster data set of natural and man-made disasters in the United States from 1990 to 2019"
author: "Catalina Canizares, Mark Macgowan, Gabriel Odom"
output: rmarkdown::html_vignette
bibliography: r-references.bib
vignette: >
  %\VignetteIndexEntry{manuscript}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
	echo = FALSE,
	message = TRUE,
	warning = TRUE
)

```

```{r setup}
library(tidyDisasters)
```

 
The Sendai Framework for Disaster Risk Reduction 2015–2030 aims to reduce disaster risk and losses in lives and physical, social, cultural, and environmental assets over the next 15 years [@undrr_2015]. For this aim to be effective, disaster risk reduction has to rely on timely and reliable information and scientific data [@Migliorini2019]. However, these data are typically dispersed and stored in different databases owned by various government agencies, research centers, private organizations, and a wide variety of stakeholders [@li2019;@Faiella2022]. 

The highly dispersed data makes it difficult for researchers to discover and use all relevant data for modeling and predicting future disasters and underlying risk factors for vulnerable territories [@li2019]. This data gap is not new; this discussion has been present in literature since 1945 [@Faiella2022], the Disaster Risk Management Knowledge Centre of the E.U. Commission has published several reports that highlight the limitations, inconsistencies, and gaps of current datasets [@Faiella2022], and the academic community has constantly advocated for the importance of the interconnection of datasets, given that they come from different scientific disciplines (hydrology, meteorology, climate) [@li2019].

Efforts have been made to address this gap. For example, the Linked Open Data for Global Disaster Risk Research (LOGD), approved by the Working Group of the Committee on Data for Science and Technology (CODATA) (Taipei, 2012), is a task force dedicated to reducing the barriers that researchers face due to the limited interconnection of disaster-related data [@codata]. The first outcome from this group is a whitepaper identifying the data gap and giving specific recommendations to the scientific community and governmental organizations to advance the state of disaster data for a more effective way of mitigating and reducing the impact of disasters [@li2019]. Another initiative is The Loss Data Enhancement for Disaster Risk Reduction and Climate Change Adaptation project (LODE), which has contributed to the data science in disaster risk by analyzing the current weaknesses in post-disaster damage and loss data management and has developed an information system for improved data collection, storage and retrieval to serve multiple applications for mitigation and preparedness [@Faiella2022].

 
In order to support these initiatives and to keep advocating for the need for open data and data interconnectivity that is accessible and usable by researchers, decision-makers, and the public, to fully understand the cause and impact of disaster events without getting lost in the "sea" of distributed and disparate data, there is a need for efforts to merge multiple information resources into a single queryable database.

To date, large amounts of disaster-related data exist for the U.S. territory, such as the Emergency Events Database (EM-DAT) maintained by the Centre for Research on the Epidemiology of Disasters (Belgium); the Global Terrorism Database (GTD) by the National Consortium for the Study of Terrorism and Responses to Terrorism (United States of America); and FEMA's National Emergency Management Information System (United States of America), that administers the FEMA Disaster Declarations Summary.
 
Taking into account the lack of interconnection between large data sets and the need for quality information for effective monitoring of the Sendai Framework and the development of evidence-based interventions, this package intended to merge the  EM-DAT, GTD, and FEMA data sets to produce a set of relational databases in a fifth standard form that unites information from three sources that complement each other. For example, whereas FEMA reports the county-level location of a natural event, EMDAT estimates the number of killed and wounded by that natural event, and the GTD contains the terrorism events. 

The merge process was based on data equivalence criteria, meaning that the reported natural disasters and terrorism events were matched at the state and county level and by the reported date or range of dates. Data cleaning for each data set was needed to reach the homogenization process using the Tidyverse package [@Wickham2019] for R Studio.

Spelling corrections for states and counties had to be performed; date formats had to be homogenized between the data sets to match the year, month, day, and format; imputation for end dates was conducted for fire incidents where no end date was reported by calculating the average of days fires lasted since 1991 according to FEMA; separation of variables into columns was performed, especially for the EM-DAT, given that in one single cell the names of country, states or county where reported. Finally, a mapping of U.S. states to counties and imputations was performed. Once the data sets were cleaned, creating a primary key that could unite the data sets was followed. This key contains the beginning year, month, state, and a ticker to differentiate events with the same Y-M-State. 

Additionally, the harmonization considered the lack of consistency between the three data sets in determining the reported event type. The United Nations Office has identified this gap for Disaster Risk Reduction, and to reduce this gap, a technical report of hazard definition and classification has been published [@UNDDR]. The report compiles a list of 302 hazards grouped according to eight clusters: meteorological and hydrological hazards, extraterrestrial hazards, geohazards, environmental hazards, chemical hazards, biological hazards, technological hazards, and societal hazards [@UNDDR]. The report states that this hazard list is considered the most useful at present [@UNDDR]. Taking this into account, this package has adopted this classification by converting the original type of event reported by EM-DAT, GTD, and FEMA and has adapted its name to the Hazard Cluster following annex 6 of the report, where the equivalences for each possible disaster is presented. The equivalence adapted was given in discussions between the package's authors and corresponded to a subjective opinion. The original hazard classification table given by the U.N. report is also part of the package to be used in case of any disagreement with the subjective classification given by the authors. 

Finally, the cleaned and harmonized data sets can be used to relate mass casualties to their short and long-term consequences beyond the damage itself. Therefore, relating mental and physical health consequences.



```{r}
 


```


