---
title: "manuscript"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{manuscript}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tidyDisasters)
```

 
Global and national datasets reporting mass casualty events are the foundation for natural disaster and terrorism research. Several global mass casualty datasets have been developed by different groups, such as the Centre for Research on the Epidemiology of Disasters (Belgium), the National Consortium for the Study of Terrorism and Responses to Terrorism (United States of America), and FEMA's National Emergency Management Information System (United States of America). This study created a new dataset by merging the data sets maintained by the groups mentioned above: the Emergency Events Database (EM-DAT), the Global Terrorism Database (GTD), and the FEMA Disaster Declarations Summary. The merge was conducted for disasters in the U.S territory and reported three or more human casualties. The merge process was based on data equivalence criteria, meaning that natural disasters and terrorism events reported were matched on state and county level and by the reported date or range of dates. To reach the homogenization process, data cleaning for each data set was needed. Using the Tidyverse package for R Studio, spelling corrections, date formats, separating variables into columns, mapping of U.S states to counties, and imputations were performed. The result is a queryable data set that unites information from three sources that complement each other. Whereas FEMA reports the county-level location of a natural event, EMDAT estimates the number of killed and wounded of that natural event, and the GTD contains the terrorism events. This new data set can be used to relate mass casualties to their short and long-term consequences beyond the damage itself. Therefore, relating mental and physical health consequences and economic and political effects, or data that is of interest, can be matched to this new data set by location (state or county level) and date. 

