---
title: "manuscript"
output: rmarkdown::html_vignette
bibliography: r-references.bib
vignette: >
  %\VignetteIndexEntry{manuscript}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
	echo = FALSE,
	message = TRUE,
	warning = TRUE
)

```

```{r setup}
library(tidyDisasters)
```

 
Global and national datasets reporting mass casualty events are the foundation for natural disaster and terrorism research. Several global mass casualty datasets have been developed by different groups, such as the Centre for Research on the Epidemiology of Disasters (Belgium), which maintains the Emergency Events Database (EM-DAT); the National Consortium for the Study of Terrorism and Responses to Terrorism (United States of America), that hold the Global Terrorism Database (GTD), and FEMA's National Emergency Management Information System (United States of America), that administers the FEMA Disaster Declarations Summary.

Using the data mentioned above, several studies relating different phenomena to disasters are available. For example, a recent report using the EM-DAT found that the trend in the occurrence of natural disasters was positively correlated with trends in ischemic heart disease mortality among females [@Huang2021]. Another current publication using the GTD constructed five multiclass classification models for the prediction of terrorist organizations that perpetrated terrorist attacks [@Xi2021]; another study using FEMA data found that the current president is more likely to grant a governor's request to declare a major disaster when facing reelection, especially in states with a larger number of electoral votes [@Husted2014].

This type of research is valuable and generates knowledge regarding the impact of disasters on humankind and society and are tools to develop evidence-based prevention and intervention programs. However, research using each data set may fail to answer questions related to the United States geography, dates, and casualties. Therefore, this package intended to merge the  EM-DAT, GTD, and FEMA data sets to produce a set of relational databases in a fifth normal form  that unites information from three sources that complement each other. Whereas FEMA reports the county-level location of a natural event, EMDAT estimates the number of killed and wounded by that natural event, and the GTD contains the terrorism events. 

The merge process was based on data equivalence criteria, meaning that the reported natural disasters and terrorism events were matched at the state and county level and by the reported date or range of dates. Data cleaning for each data set was needed to reach the homogenization process. Using the Tidyverse package [@Wickham2019] for R Studio.

Spelling corrections for states and counties had to be performed; date formats had to be homogenized between the data sets to match the year, month, day, format; imputation for end dates were conducted for fire incidents where no end date was reported by calculating the average of days fires lasted since 1991 according to FEMA; separation of variables into columns was performed, especially for the EM-DAT, given that in one single cell the names of country, states or county where reported. Finally, a mapping of U.S states to counties and imputations was performed. Once the data sets were cleaned, creating a primary key that could unite the data sets was followed. This key contains the beginning year, month, state, and a ticker to differentiate events with the same Y-M-State. 

This new data set can be used to relate mass casualties to their short and long-term consequences beyond the damage itself. Therefore, relating mental and physical health consequences and economic and political effects, or data of interest, can be matched to this new data set by location (state or county level) and date.

Questions that may be answered with this new data set include

```{r}
 


```


