---
title: "manuscript"
output: rmarkdown::html_vignette
bibliography: r-references.bib
vignette: >
  %\VignetteIndexEntry{manuscript}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
	echo = FALSE,
	message = TRUE,
	warning = TRUE
)

```

```{r setup}
library(tidyDisasters)
```

 

Research on disaster risk relies heavily on scientific data. To date, large amounts of disaster-related data exist, such as the Emergency Events Database (EM-DAT) maintained by the Centre for Research on the Epidemiology of Disasters (Belgium); the Global Terrorism Database (GTD) by the National Consortium for the Study of Terrorism and Responses to Terrorism (United States of America); and FEMA's National Emergency Management Information System (United States of America), that administers the FEMA Disaster Declarations Summary. 

However, these data are typically dispersed, stored in different data bases owned by various government agencies, research centers, private organizations and a wide variety of stakeholders [@li2019]. The highly dispersed data makes it difficult for  researchers to discover and use all relevant data needed for a certain study. In fact, when they are able to find the appropriate data it sometimes come from different data sets that lack of single identifier which precludes researchers to successfully select, transfer, interpret and use the data [@li2019]. 

The need for open data and data interconnectivity, i.e. data which are accessible and usable by researchers, decision-makers and the public, is most critical in the area of disaster risk research, management and mitigation.
In-depth scientific analysis of disaster data requires management and acquisition patterns that support interconnection of dispersed scientific data related to disaster risk assessment and reduction. Disaster risk researchers face perhaps a greater challenge to find relevant datasets in a “sea” of distributed and disparate data resources. Therefore, gaps in data infrastructure, data sharing policies and data use governance should be addressed to unleash the potential of disaster risk research in helping regions, especially the developing countries, to improve risk assessment, reduction and management.


This type of research is valuable and generates knowledge regarding the impact of disasters on humankind and society and are tools to develop evidence-based prevention and intervention programs. Also, they are in line with the The Sendai Framework for Disaster Risk Reduction 2015–2030 aim “To strengthen technical and scientific capacity to capitalize on and consolidate existing knowledge and to develop and apply methodologies and models to assess disaster risks, vulnerabilities and exposure to all hazards” (UNISDR 2015). However, this studys are subject to the 



Task group of LODGD will study the mechanism for connecting such data to enable easier and faster discovery and access, and significantly reduce the barriers that researchers are facing today due to limited interconnection of various disaster-related data http://www.codata.info/taskgroups/TGdisasterrisk/




Therefore, this package intended to merge the  EM-DAT, GTD, and FEMA data sets to produce a set of relational databases in a fifth normal form  that unites information from three sources that complement each other. Whereas FEMA reports the county-level location of a natural event, EMDAT estimates the number of killed and wounded by that natural event, and the GTD contains the terrorism events. 

The merge process was based on data equivalence criteria, meaning that the reported natural disasters and terrorism events were matched at the state and county level and by the reported date or range of dates. Data cleaning for each data set was needed to reach the homogenization process. Using the Tidyverse package [@Wickham2019] for R Studio.

Spelling corrections for states and counties had to be performed; date formats had to be homogenized between the data sets to match the year, month, day, format; imputation for end dates were conducted for fire incidents where no end date was reported by calculating the average of days fires lasted since 1991 according to FEMA; separation of variables into columns was performed, especially for the EM-DAT, given that in one single cell the names of country, states or county where reported. Finally, a mapping of U.S states to counties and imputations was performed. Once the data sets were cleaned, creating a primary key that could unite the data sets was followed. This key contains the beginning year, month, state, and a ticker to differentiate events with the same Y-M-State. 

Additionally, the 

This new data set can be used to relate mass casualties to their short and long-term consequences beyond the damage itself. Therefore, relating mental and physical health consequences and economic and political effects, or data of interest, can be matched to this new data set by location (state or county level) and date.

Questions that may be answered with this new data set include

```{r}
 


```


