---
title: "`tidyDisasters` R Package: A harmonized disaster data set of natural and man-made disasters in the United States from 1990 to 2020"
author: "Catalina Canizares, Mark Macgowan, Gabriel Odom"
output: rmarkdown::html_vignette
bibliography: r-references.bib
vignette: >
  %\VignetteIndexEntry{`tidyDisasters` R Package: A harmonized disaster data set of natural and man-made disasters in the United States from 1990 to 2020}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)

```

```{r setup, include=FALSE}
library(tidyDisasters)
library(lubridate)
library(gghighlight)
library(tidyverse)
```

## Introduction 

The Sendai Framework for Disaster Risk Reduction 2015–2030 aims to reduce disaster risk and losses in lives and physical, social, cultural, and environmental assets over the next 15 years [@undrr_2015]. For this aim to be effective, disaster risk reduction has to rely on timely and reliable information and scientific data [@Migliorini2019]. However, these data are typically dispersed, unstructured and stored in different databases owned by various government agencies, research centers, private organizations, and a wide variety of stakeholders [@li2019;@Faiella2022; @Ortmann; @Nunavath2017]. 

The highly dispersed data makes it difficult for researchers to discover and use all relevant data for modeling and predicting future disasters and underlying risk factors for vulnerable territories [@li2019]. This data gap is not new; this discussion has been present in literature since 1945 [@Faiella2022], the Disaster Risk Management Knowledge Centre of the E.U. Commission has published several reports that highlight the limitations, inconsistencies, and gaps of current datasets [@Faiella2022], and the academic community has constantly advocated for the importance of the interconnection of datasets, given that they come from different scientific disciplines (hydrology, meteorology, climate) [@li2019].

Efforts have been made to address this gap. For example, the Linked Open Data for Global Disaster Risk Research (LOGD), approved by the Working Group of the Committee on Data for Science and Technology (CODATA) (Taipei, 2012), is a task force dedicated to reducing the barriers that researchers face due to the limited interconnection of disaster-related data [@codata]. The first outcome from this group is a whitepaper identifying the data gap and giving specific recommendations to the scientific community and governmental organizations to advance the state of disaster data for a more effective way of mitigating and reducing the impact of disasters [@li2019]. Another initiative is The Loss Data Enhancement for Disaster Risk Reduction and Climate Change Adaptation project (LODE), which has contributed to the data science in disaster risk by analyzing the current weaknesses in post-disaster damage and loss data management and has developed an information system for improved data collection, storage and retrieval to serve multiple applications for mitigation and preparedness [@Faiella2022].

In order to support these initiatives and to keep advocating for the need for open data and data interconnectivity that is accessible and usable by researchers, decision-makers, and the public, to fully understand the cause and impact of disaster events without getting lost in the "sea" of distributed and disparate data [@li2019], there is a need for efforts to merge multiple information resources into a single queryable database.

To date, large amounts of disaster-related data exist for the U.S. territory, such as the Emergency Events Database (EM-DAT) maintained by the Centre for Research on the Epidemiology of Disasters (Belgium); the Global Terrorism Database (GTD) by the National Consortium for the Study of Terrorism and Responses to Terrorism (United States of America); and Federal Emergency Management Agency (FEMA) National Emergency Management Information System (United States of America), that administers the FEMA Disaster Declarations Summary.
 
Taking into account the lack of interconnection between large data sets and the need for quality information for effective monitoring of the Sendai Framework and the development of evidence-based interventions, this package intended to merge the  EM-DAT, GTD, and FEMA data sets to produce a set of relational databases in a fifth normal form [@kent1983] that unites information from three sources that complement each other.  For example, whereas FEMA reports the county-level location of a natural event, EMDAT estimates the number of killed and wounded by that natural event, and the GTD contains the terrorism events. 

## Method

The merge was conducted for disasters since January 1990 to 2020 in the 50 U.S. states (plus the District of Columbia and Puerto Rico) and for which three or more human casualties were reported [@bill]. The merge process was based on data equivalence criteria, meaning that the reported natural disasters and terrorism events were matched at the state and county level and by the reported date or range of dates. Data cleaning for each data set was needed to reach the homogenization process using the Tidyverse package [@Wickham2019] for R Studio.

Spelling corrections for states and counties had to be performed (see table `disastLocation_df`); date formats had to be homogenized between the data sets to match the year, month, day, and format (see table `distastDates_df`); imputation for end dates was conducted for fire incidents where no end date was reported by calculating the average of days fires lasted since 1991 according to FEMA; separation of variables into columns was performed, especially for the EM-DAT, given that in one single cell the names of country, states or county where reported. Finally, a mapping of U.S. states to counties and imputations was performed. Once the data sets were cleaned, creating a primary key that could unite the data sets was followed (see table `allKeys_df`). This key contains the beginning year, month, state, and a ticker to differentiate events with the same year, month and state.  

Additionally, the harmonization considered the lack of consistency between the three data sets in determining the reported event type. The United Nations Office has identified this gap for Disaster Risk Reduction, and to reduce this gap, a technical report of hazard definition and classification has been published [@UNDDR]. The report compiles a list of 302 hazards grouped according to eight clusters: meteorological and hydrological hazards, extraterrestrial hazards, geohazards, environmental hazards, chemical hazards, biological hazards, technological hazards, and societal hazards [@UNDDR]. The report states that this hazard list is considered the most useful at present [@UNDDR]. Taking this into account, this package has adopted this classification by converting the original type of event reported by EM-DAT (see table `emdat_hazard_cluster_df`), GTD, and FEMA (see table `fema_hazard_cluster_df`) and has adapted its name to the Hazard Cluster following annex 6 of the report, where the equivalences for each possible disaster is presented. The equivalence adapted was given in discussions between the package's authors and corresponded to the author educated opinion (see table `disastTypes_df`). The original hazard classification table given by the U.N. report is also part of the package (see table `hazard_report_df`) to be used in case of any disagreement with the subjective classification given by the authors. 

Finally, the package contains eight tables that share the same key to be efficiently joined. The tables were constructed in a fifth normal form [@kent1983] to save computer memory and are presented in table 1. 

Table 1, Date sets inside the `tidyDisasters` package

| Name of data set           | Description                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                 |
| -------------------------- | --------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------- |
| `allKeys_df`                | This data contains the original keys from the FEMA, GTD and EMDAT data set and the universal key created by the authors.                                                                                                                                                                                                                                                                                                                                                                                                    |
| `disastCasualties_df `      | This data contains the number of people killed and wounded per event.<br>It is relevant to note that the number of people killed and wounded<br>correspond to the total casualties for the whole event, it does not relate<br>to the particular number of casualties in each county or state                                                                                                                                                                                                                                |
| `disastDates_df`            | This data contains the beginning and end date of the disasters<br>registered in EMDAT and FEMA. The dates are given at the state level, this means that if different counties where affected in different dates this will not be accounted for in this data set.                                                                                                                                                                                                                                                            |
| `disastLocations_df `       | This data contains the states and counties where disasters<br> have happened across the USA.                                                                                                                                                                                                                                                                                                                                                                                                                                |
| `disastTypes_df`            | This data contains contains the type of disasters reported<br>by FEMA and EMDAT based on the classification  of the Hazard Definition and Classification Review Technical Report  published by the UN Office for Disaster Risk Reduction (2020)                                                                                                                                                                                                                                                                             |
| `emdat_hazard_cluster_df` |  This data contains contains the type of disasters reported<br> by EMDAT  and the matched hazard cluster based on the classification  of  the Hazard Definition and  Classification Review Technical Report<br>published by the UN Office for  Disaster Risk Reduction (2020)<br>Annex 6 page 72. The matching was done by Dr. Mark Macgowan and<br> Catalina Canizares following the suggestions in the report. However,<br> this classification is subjective and should be used taking into<br> account this disclaimer. |
| `fema_hazard_cluster_df`  | This data contains the type of disasters reported  by FEMA  and the matched hazard cluster based on the classification  of<br>the Hazard Definition and  Classification Review Technical Report<br>published by the UN Office for  Disaster Risk Reduction (2020)<br>Annex 6 page 72. The marching was done by Dr. Mark Macgowan and<br>Catalina Canizares following the suggestions in the report. However,<br> this classification is subjective and should be used taking into<br>account this disclaimer.               |
| `hazard_report_df`         |  This data contains the table from Annex 6 page 72 of the Hazard<br>Definition and Classification Review Technical Report published by the<br>UN Office for Disaster Risk Reduction (2020)                                                                                                                                                                                                                                                                                                                                  |

## Examples

With the data that is contained in the package it is possible to query and plot information easily. Examples are given bellow. 

### Example 1: Considers time trends and U.S geography affected
The first example explores the maximum number of counties affected by a single severe ice and snow storm. By joining the cleaned data sets of locations (`disastLocations_df)`, types of events (`disastTypes_df`)  and dates (`disastDates_df`) we were able to plot and find the 1993 "Storm of the Century" that affected almost 800 counties [@NOAA2020]  and the 1996 blizzard that affected around 500 counties and is remembered as the greatest snowstorm in terms of the amount of snow that fell [@Stachelski]. 

```{r example-1, echo=TRUE, fig.dim = c(6, 4), dev='svg'}
data("disastLocations_df")
data("disastTypes_df")
data("disastDates_df")

disastLocations_df %>%
  left_join(disastTypes_df) %>%
  left_join(disastDates_df) %>%
  mutate(Year = year(eventStart)) %>% 
  filter(incident_type %in% c("Severe Ice Storm", "Snow")) %>%
  group_by(state, county, Year) %>%
  summarise(Frozen = n() >= 1L, .groups = "keep") %>%
  group_by(Year) %>%
  summarise(Count = sum(Frozen)) %>%
  ggplot() +
  aes(x = Year, y = Count) +
  labs(
    title = "Number of Counties Affected by Snow Storms Since the 90s",
    caption = "Data from the tidyDisasters R Package",
    y = "No. Counties Frozen"
  ) +
  theme_classic() +
  geom_point() +
  gghighlight(Count > 400)


```

### Example 2: Considers time trends and Hazard Clusters 

Another example of the usage of this package is this plot that evidences the trend on  six different types of hazard clusters.

```{r example-2, echo=TRUE, fig.dim = c(8, 4), dev='svg'}

  disastTypes_df %>%
  left_join(disastDates_df) %>%
  mutate(Year = year(eventStart)) %>% 
  filter(
    hazard_cluster %in% c(
      "Precipitation-related", " Wind-related",
      "Environmental degradation (Forestry)", "Flood", "Convective-related",
      "Pressure-related"
    )
  ) %>%
  group_by(Year, hazard_cluster) %>%
  count(hazard_cluster) %>%
  ggplot() +
  aes(x = Year, y = n, color = hazard_cluster) +
  labs(
    title = "Disasters tendency since 1990",
    caption = "Data from the tidyDisasters R Package",
    y = "Count"
  ) +
  geom_point() +
  facet_wrap(~hazard_cluster, scales = "free_y") +
  theme_classic() +
  theme(legend.position = "none") 
  
  

```

### Example 3: Considers individual query of specifi event 

Finally, another example from this data show that it can be used to query individual event such as the Orlando nightclub shooting in 2016. 

```{r example-3, echo=TRUE}
data("disastCasualties_df")

 disastTypes_df %>%
  left_join(disastDates_df) %>%
  left_join(disastCasualties_df) %>% 
  left_join(disastLocations_df) %>% 
  mutate(Year = year(eventStart)) %>% 
  filter(state == "FL" & hazard_type == "SOCIETAL" & Year == 2016) %>% 
  rmarkdown::paged_table()
  

```

## Discussion

The need for integrated and queryable data in disaster management is essential for better emergency response and prevention of loss. However, the data that is currently available tends to be disperse, in different formats, heterogeneously distributed and owned by different stakeholders and agencies [@li2019;@Faiella2022; @Ortmann; @Nunavath2017]. Consequently, interventions and prevention programs are taking longer to develop given the lack of data and decisions become difficult to take given the limited information avaialble [@Nunavath2017. Therefore, this study aimed at uniting  available heterogeneously dispersed data from three different sources  together to improve information accessibility and analysis. 

Moreoever, the cleaned and harmonized data sets can be used to relate mass casualties to their short and long-term consequences beyond the damage itself, because other data sets can be easily joined if there is state and county level information. Therefore, mental and physical health consequences, economic loss, or other variables of interest can be analyzed taking into account this already cleaned and joined data. 

## References
